\section{Conclusion: beyond raw scaling}

Julia Kempe's lecture presents a sobering but constructive view of the synthetic-data age. Scaling laws were observed under assumptions about data distributions that recursive model-generated corpora can violate. The consequences include altered exponents, error floors, lost skills, recursive label bias, delayed recovery, and verifier-dependent phase transitions.

The lecture does not imply that synthetic data should be rejected. It establishes a more precise doctrine:

\begin{enumerate}[leftmargin=*]
  \item synthetic data are most useful when real data are scarce or when they target otherwise inaccessible regimes;
  \item synthetic volume cannot recover support that the generator never produces;
  \item recursive pseudo-labels require independent corrective signals;
  \item real data, accumulation, and verification can change the asymptotic outcome;
  \item data curation becomes part of model architecture and risk governance.
\end{enumerate}

For MMALS, the deepest translation is that an adaptive system must preserve the conditions of future surprise. A route should not become dominant merely because it generated most of the replay. A memory should not become true because several system generations repeated it. A rare host should not be maintained for diversity theatre, but neither should it disappear before its causal value is tested.

The target architecture is therefore:
\[
\boxed{
\text{local simplicity}
+\text{global functional diversity}
+\text{independent verification}
+\text{lineage-aware memory}
}
\]

For STRAT-Q and probabilistic MTP, evidence must be discounted by dependence, lineage, freshness, oracle uncertainty, and missing-tail risk. The number of tests is not the number of independent proofs. A cryptographically verifiable credential is an authenticity layer, not an epistemic conclusion.

For engineering and scientific machine learning, simulation and generation remain powerful. Their role is to propose, explore, densify, and stress. Real measurements and validated constraints remain necessary to correct model-form error.

\begin{keyidea}
The next era is not literally ``beyond scaling'' in the sense that scale no longer matters. It is beyond the belief that scale is sufficient. What must scale is verified, diverse, fresh, decision-relevant functional support.
\end{keyidea}
